The Galactic Midplane Is Not a Plane: Implications for Dynamical Analysis with {\em Gaia} Data and Beyond

### 2019-04-23 paper/ms.tex
Add ref to Kamdar paper
Update title, subsection titles

\title{The Galactic Midplane is Not a Plane: Implications for dynamical analysis with {\em Gaia} data and beyond}

% \correspondingauthor{Angus Beane}
%mm [will this address go away during the review process?.  Do you
%want to give a "current address" as a second footnote?]
\email{abeane@sas.upenn.edu}

@@

\subsection{Cartoon effect of midplane error} \label{ssec:cartoon}
We present a cartoon approximation in Figure~\ref{fig:cartoon} to show how an
inaccurate determination of the midplane leads to a phase-dependence of the
actions calculated from an orbit specified by a phase-space point and an
assumed potential model. The $y$-axis corresponds to the vertical height of
the orbit and the $x$-axis orbital phase. The solid gray line indicates the
``true'' orbit of the star; that is, its orbit as it oscillates around the
true midplane. The dashed gray line is an erroneous midplane assigned to the
potential model which an observer uses to integrate the orbit of the star and
hence estimate its actions. The model potential is otherwise identical to the
one in which the star is actually moving.

@@

\subsection{Numerical methods} \label{ssec:action_comp}
We now move on to quantifying the argument made in Section~\ref{ssec:cartoon}.
We compute actions as in our previous work \citep{2018ApJ...867...31B}, using
the code \texttt{gala} v0.3 to perform orbit integrations and conversion to
action space \citep{2017JOSS....2..388P,Price-Whelan:2018}. To compute actions
we use the torus-mapping technique first presented by
\citet{1990MNRAS.244..634M} and adapted by \citet{2014MNRAS.441.3284S} to
calculate actions for an orbital time-series starting from a phase-space
position $(x, v)$ and integrated in a potential $\Phi$. We use the default
\texttt{MWPotential} as our potential, which is based on the Milky Way
potential available in \texttt{galpy} \citep{2015ApJS..216...29B}. This
potential includes a Hernquist bulge and nucleus \citep{1990ApJ...356..359H},
a Miyamoto-Nagai disk \citep{1975PASJ...27..533M}, and an NFW halo
\citep{1997ApJ...490..493N}, and is fit to empirically match some
observations. We use the Dormand-Prince 8(5,3) integration scheme
\citep{Dormand80:integrator} with a timestep of $1\,\Myr$ and integrate for
$5\,\Gyr$, corresponding to $\sim 20$ orbits for a Sun-like star.

@@

\subsection{Quantification of the midplane effect} \label{ssec:quant}
We quantify how a systematic error in the Galactocentric coordinate system
induces phase-dependence in the {\em observed} actions. We will consider three
orbits in the model potential described in Section~\ref{ssec:action_comp} that
are typical of stars in the stellar thin disk, stellar thick disk, and the
stellar halo. We summarize their initial positions in phase space and the
actions computed by integrating their orbits in the correct potential in
Table~\ref{tab:orbits}. Each orbit, integrated without systematic coordinate
errors, is plotted in Appendix~\ref{app:orbits}. We will refer to these orbits
by their names (thin-disk, thick-disk, halo) henceforth.

@@

% \section{Azimuthal midplane variations at the Solar Circle: estimates from
% simulation} 
\section{Coordinate System Offsets II. Azimuthal Midplane Variations in
Simulations}
\label{sec:local_fire}

@@

\subsection{Description of FIRE Simulation} \label{ssec:cosmozoom}
We briefly describe the cosmological-hydrodynamical zoom-in simulations used
in this work. Cosmological zoom-ins allow one to simulate a selected region at
high resolution embedded in a low-resolution cosmological background
\citep[e.g.][]{1993ApJ...412..455K,2014MNRAS.437.1894O}, to which can be added
a hydrodynamical simulation of the baryonic component with recipes for star
formation and feedback. The FIRE collaboration has simulated a number of Milky
Way-mass galaxies using this technique as part of the {\em Latte} suite of
FIRE-2 simulations \citep{2016ApJ...827L..23W,2018MNRAS.481.4133G}. For this
work we focus on the three zoom-ins considered in \citet{2018arXiv180610564S},
which show broad agreement of many of their global properties with
observations of the Milky Way. The $\z=0$ snapshots\footnote{In this work, to
avoid confusion with the vertical height $z$, we refer to cosmological
redshift as $\z$.} of these three simulations, called \mi{}, \mf{}, and \mm{}
for shorthand, are publicly available alongside associated mock {\em Gaia} DR2
catalogues generated from them.\footnote{\url{http://ananke.hub.yt}}

@@

\subsection{The size of the solar neighborhood} \label{sssec:neighborhood}
Given a local coordinate system, the amplitude and characteristic length scale
of the density fluctuations in the Galactic disk will determine how far this
coordinate system can be extended before introducing relevant systematic
errors in the computation of actions for thin-disk stars. We consider a $10\%$
error to be a criterion for the region around the Sun where the analysis of
the action-space distribution of stars on quasi-circular orbits, and therefore
the epicyclic approximation, can be expected to be valid. This leads to a
somewhat dynamically motivated definition of the ``solar neighborhood.'' Since
the motivation may be different for defining the solar neighborhood in, e.g.,
studies of the chemical distribution of the Galaxy, we propose the phrase
``dynamical solar neighborhood.'' We also point out that our use of a $10\%$
error is arbitrary.

@@

\subsection{Consequences for disrupted cluster reconstruction}
\label{sssec:reconstruction}

@@

\item This work emphasizes the importance of combining chemistry and dynamics.
Chemical tagging \citep{2002ARA&A..40..487F} and dynamical tagging must
complement and confirm each other. This was also recently pointed out by
\Gus{someone}.

>>>>

\title{The Galactic Midplane Is Not a Plane: Implications for Dynamical
Analysis with {\em Gaia} Data and Beyond}

\correspondingauthor{Angus Beane}
\email{abeane@sas.upenn.edu}

@@

\subsection{Cartoon Effect of Midplane Error} \label{ssec:cartoon}
We present a cartoon approximation in Figure~\ref{fig:cartoon} to show how an
inaccurate determination of the midplane leads to a phase-dependence of the
actions calculated from an orbit specified by a phase-space point and an
assumed potential model. The $y$-axis corresponds to the vertical height of
the orbit and the $x$-axis orbital phase. The solid gray line indicates the
``true'' orbit of the star; that is, its orbit as it oscillates around the
true midplane. The dashed gray line is an erroneous midplane assigned to the
potential model which an observer uses to integrate the orbit of the star and
hence estimate its actions. The model potential is otherwise identical to the
one in which the star is actually moving.

@@

\subsection{Numerical Methods} \label{ssec:action_comp}
We now move on to quantifying the argument made in Section~\ref{ssec:cartoon}.
We compute actions as in our previous work \citep{2018ApJ...867...31B}, using
the code \texttt{gala} v0.3 to perform orbit integrations and conversion to
action space \citep{2017JOSS....2..388P,Price-Whelan:2018}. To compute actions
we use the torus-mapping technique first presented by
\citet{1990MNRAS.244..634M} and adapted by \citet{2014MNRAS.441.3284S} to
calculate actions for an orbital time-series starting from a phase-space
position $(x, v)$ and integrated in a potential $\Phi$. We use the default
\texttt{MWPotential} as our potential, which is based on the Milky Way
potential available in \texttt{galpy} \citep{2015ApJS..216...29B}. This
potential includes a Hernquist bulge and nucleus \citep{1990ApJ...356..359H},
a Miyamoto-Nagai disk \citep{1975PASJ...27..533M}, and an NFW halo
\citep{1997ApJ...490..493N}, and is fit to empirically match some
observations. We use the Dormand-Prince 8(5,3) integration scheme
\citep{Dormand80:integrator} with a timestep of $1\,\Myr$ and integrate for
$5\,\Gyr$, corresponding to $\sim 20$ orbits for a Sun-like star.

@@

\subsection{Quantification of the Midplane Effect} \label{ssec:quant}
We quantify how a systematic error in the Galactocentric coordinate system
induces phase-dependence in the {\em observed} actions. We will consider three
orbits in the model potential described in Section~\ref{ssec:action_comp} that
are typical of stars in the stellar thin disk, stellar thick disk, and the
stellar halo. We summarize their initial positions in phase space and the
actions computed by integrating their orbits in the correct potential in
Table~\ref{tab:orbits}. Each orbit, integrated without systematic coordinate
errors, is plotted in Appendix~\ref{app:orbits}. We will refer to these orbits
by their names (thin-disk, thick-disk, halo) henceforth.

@@

\section{Coordinate System Offsets II. Azimuthal Midplane Variations in
Simulations}
\label{sec:local_fire}

@@

\subsection{Description of FIRE Simulations} \label{ssec:cosmozoom}
We briefly describe the cosmological-hydrodynamical zoom-in simulations used
in this work. Cosmological zoom-ins allow one to simulate a selected region at
high resolution embedded in a low-resolution cosmological background
\citep[e.g.][]{1993ApJ...412..455K,2014MNRAS.437.1894O}, to which can be added
a hydrodynamical simulation of the baryonic component with recipes for star
formation and feedback. The FIRE collaboration has simulated a number of Milky
Way-mass galaxies using this technique as part of the {\em Latte} suite of
FIRE-2 simulations \citep{2016ApJ...827L..23W,2018MNRAS.481.4133G}. For this
work we focus on the three zoom-ins considered in \citet{2018arXiv180610564S},
which show broad agreement of many of their global properties with
observations of the Milky Way. The $\z=0$ snapshots\footnote{In this work, to
avoid confusion with the vertical height $z$, we refer to cosmological
redshift as $\z$.} of these three simulations, called \mi{}, \mf{}, and \mm{}
for shorthand, are publicly available alongside associated mock {\em Gaia} DR2
catalogues generated from them.\footnote{\url{http://ananke.hub.yt}}

@@

\subsection{The Size of the Solar Neighborhood} \label{sssec:neighborhood}
Given a local coordinate system, the amplitude and characteristic length scale
of the density fluctuations in the Galactic disk will determine how far this
coordinate system can be extended before introducing relevant systematic
errors in the computation of actions for thin-disk stars. We consider a $10\%$
error to be a criterion for the region around the Sun where the analysis of
the action-space distribution of stars on quasi-circular orbits, and therefore
the epicyclic approximation, can be expected to be valid. This leads to a
somewhat dynamically motivated definition of the ``solar neighborhood.'' Since
the motivation may be different for defining the solar neighborhood in, e.g.,
studies of the chemical distribution of the Galaxy, we propose the phrase
``dynamical solar neighborhood.'' We also point out that our use of a $10\%$
error is arbitrary.

@@

\subsection{Disrupted Cluster Reconstruction}
\label{sssec:reconstruction}

@@

\item This work emphasizes the importance of combining chemistry and dynamics.
Chemical tagging \citep{2002ARA&A..40..487F} and dynamical tagging must
complement and confirm each other. The importance of combining chemistry and
dynamics for Galactic archaeology was recently pointed out by
\citet{2019arXiv190210719K}.
